---
title: "Time-Series-1"
author: "Amey Joshi"
date: "13/06/2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("astsa")
set.seed(24041996)
```

## What is a time series?
- Time series is a collection of random variables indexed according to the order
they were obtained in time.
- A stochastic process is to a time series what a random variable is to a number.
The latter is a realisation of the former in an experiment. Repeating the 
experiment will result in a different time series.
- Some examples:
  - (Finance) asset prices
  - (Economics) GDP, unemployment, inflation
  - (Epidemiology) # cases of infection
  - (Demography) birth and death rates
  - (Medicine) blood sugar levels in a day, ECG, EEG, fMRI
  - (Physics) wind speeds, ocean temperature, solar activity
- Time series analysis is closely related to signal processing.

## How do time series look? J&J quarterly earnings - 1

```{r}
plot(
  jj,
  type = "o",
  xlab = "year",
  ylab = "quarterly earnings",
  main = "Johnson&Johnson",
  cex = 0.1
)
```

## How do time series look? J&J quarterly earnings - 2
Key observations:

- Earnings are rising
- But not monotonically, there are clearly discernable cycles.
- As time goes by, the drop in earnings gets more severe.

## How does a time series look? Sensex - 1

```{r}
sensex_master <- read.csv("sensex.csv")
plot(sensex_master$Close, cex=0.1, ylab = "Sensex", main = "Since 3-Jan-2000")
```

## How does a time series look? Sensex - 2
Key observations:

- In general the index is rising, except for a few downturns.
- We can see turbulence around 2000 and 4000 but otherwise it 'appears smooth'.

## How does a time series look? Sensex daily returns - 1

```{r}
delta_sensex <- diff(sensex_master$Close)
daily_returns <-
  delta_sensex / sensex_master$Close[seq_along(delta_sensex)]
plot(
  daily_returns,
  cex = 0.1,
  type = "o",
  ylab = "Daily return",
  main = "Since 3-Jan-2000"
)
```

## How does a time series look? Sensex daily returns - 2
Key observations:

- Unlike the sensex series, the returns appear to have periodic bouts of high
variation.
- If we were to measure the variance over a moving window we will see a sharp
change in its values although the series seems to have a mean of zero.

## How does a time series look? Ecological example - 1
This is the plot of southern oscillation index and the number of new fish in the
Pacific. The southern oscillation index (SOI) is the difference between air 
pressures at Tahiti and Darwin in Australia. Prolonged periods of negative 
(positive) SOI values coincide with abnormally warm (cold) ocean waters across 
the eastern tropical Pacific typical of El Niño (La Niña) episodes.

## How does a time series look? Ecological example - 2
```{r}
par(mfrow = c(2, 1), oma = c(5,4,0,0)+0.1,mar = c(0,0,2,2))
plot(soi, ylab = "", xlab = "", main = "SOI (top) and #fish (bottom)")
plot(rec, ylab = "", xlab = "")
```

## How does a time series look? Ecological example - 3
Key observations:

- Both series show a periodicity
- The variation of SOI is small until the series jumps from one level to another.
- Periodic phenomena are usually studied in the frequency domain.
- The fish population seems to be a response to the SOI. Such causal relations
are studied using transfer functions.

## Why time series differs from other samples? - 1
- Unlike the usual samples whose members are selected independently of each other
the observations in a time series are correlated.
- Sensex ranges from 2600.12 to 40267.62 since 3-Jan-2000. If we choose a random
sample of numbers in this range and plot them we will observe the points to be
spread uniformly across the plot area. We also plot the sensex to show that a 
time series sample is far more orderly than the random sample.

## Why time series differs from other samples? - 2
```{r}
random_sample <-
  runif(
    n = nrow(sensex_master),
    min = min(sensex_master$Close),
    max = max(sensex_master$Close)
  )
plot(
  random_sample,
  cex = 0.1,
  type = "p",
  ylab = "random sample",
  main = "Time series versus random sample"
)
points(sensex_master$Close, cex = 0.1, col = "red")
legend(
  "topleft",
  legend = c("sample", "sensex"),
  col = c("black", "red"),
  pch = 1,
  cex = .8,
  pt.cex = .1
)
```

## Continuous and discrete time series
- It is common to join the points of a time series to make it appear as if it
were recorded continuously.
- Some series are genuinely discrete.
- A continuous time series is sampled at a certain frequency to obtain a discrete
series.
  - The sampling has to be sufficiently frequent to get a faithful representation
  of the underlying series.
  - Recall Nyquist criterion.
  
## A simulated time series - White noise
```{r}
plot(rnorm(1000), ylab="y[n]", main = "White noise", cex = 0.1, type = "o")
```

## A simulated time series - Random walk
```{r}
plot(cumsum(sample(x = c(-1, 1), replace = TRUE, size = 1000)), 
     cex = .1, type = "o", main = "Random walk", ylab = "Displacement")
```

## A simulated time series - Random walk with drift
```{r}
plot(
  cumsum(0.05 + sample(
    x = c(-1, 1),
    replace = TRUE,
    size = 1000
  )),
  cex = .1,
  type = "o",
  main = "Random walk with drift",
  ylab = "Displacement"
)
```

## Statistical description of time series - 1
- A full statistical description of a time series is its joint probability
distribution $$
F_{t_1\cdots t_n}(c_1, \ldots, c_n) = \mathrm{P}(x_{t_1} \le c_1, \ldots x_{t_n} \le c_n)
$$
- A marginal distribution at time $t$ is $F_t(c) = \mathrm{P}(x_t \le c)$. The
corresponding density function is
$$
f_t(x) = \frac{\partial F_t}{\partial x}
$$
- The mean function is defined as
$$
\mu_{xt} = \mathrm{E}(x_t) = \int x_t f_t(x) dx
$$
- Note that the mean function depends on $t$. Every $x_t$ will, in general, have 
a different expectation.

## Statistical description of time series - 2
- Recall the formula for covariance between random variables $X$ and $Y$, 
$$
\text{cov}(X, Y) = \mathrm{E}\left[(X - \mathrm{E}(X))(Y - \mathrm{E}(Y))\right]
$$
- The same idea is applied to the time series to get 'auto'-covariance
$$
\gamma_x(s, t) = \mathrm{E}\left[(x_s - \mu_s)(x_t - \mu_t)\right]
$$

- The autocovariance function is the covariance between two 'data points' of a 
time series.

## Statistical description of time series - 3
- Analgous to the correlation between random variables, the auto-correlation 
function (ACF) of a time series is
$$
\rho_x(s, t) = \frac{\gamma_x(s, t)}{\sigma_x(s)\sigma_x(t)}
$$
- It follows immediately that $\rho_x(s, s) = 1$.
- ACF of a smooth time series will be non-zero even if $s$ and $t$ are far from
each other. On the other hand, that of a choppy series will decay to zero very 
soon. In the extreme case of white noise
$$
\rho(s, t) = \begin{cases}
1 & s =t \\
0 & s \ne t
\end{cases}
$$

## Statistical description of time series - 4
- ACF answers the question, "Can we predict $x_t$ if we know $x_s$ using a 
linear relation?".
- Supposing $x_t = \beta_0 + \beta_1 x_s$. That is, $x_t$ is prefectly predicted
using $x_s$. Then $\rho(s t) = \pm 1$.
- We can analogously define the cross-covariance function between two time
series $x_y$ and $y_t$ as
$$
\gamma_{xy}(s, t) = \mathrm{E}\left[(x_s - \mu_x)(y_t - \mu_t)\right]
$$
- The cross-correlation function is
$$
\rho_{xy}(s, t) = \frac{\gamma_{xy}(s, t)}{\sigma_x(s)\sigma_y(t)}
$$
