---
title: "Time-Series-2"
author: "Amey Joshi"
date: "16/06/2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("astsa")
set.seed(24041996)
```

## Strictly stationary time series - 1
A time series $\{x_t\}$ is called strictly stationary if
$$
\mathrm{P}(x_{t_1} < c_1 \ldots x_{t_k} < c_k) = 
\mathrm{P}(x_{t_{1 + h}} < c_1 \ldots x_{t_{k + h}} < c_k)
$$
for all $k \ge 1$, all $t_1, \ldots, t_k$ and all time shifts $h \in \boldsymbol{Z}$

What does this mean?

- Choose any $k$ consecutive points of the time series. For any $h \in \boldsymbol{Z}$ 
select points that are a distance $h$ away from the previous set. If the time
series is stationary then the statistical properties of the two sets of points
are identical.
- By statistical properties we mean the joint distribution. Equivalently, all
moments of the sets are identical.
- In particular, $\rho_x(s, t) = \rho_x(s + h, t + h)$ for all $h \in 
\boldsymbol{Z}$.

## Strictly stationary time series - 2
- The property of being stationary is \emph{very} strong.
- Although one can construct a stationary time series one can never confirm that
a given time series is stationary.
  - Recall that the joint probability distribution has to be identical for all
  $h \in \boldsymbol{Z}$.
- Some time series can be classified as non-stationary by just plotting them.
  - The rising trend of the sensex.
  - A random walk with drift.
- On the other hand, some time series 'appear to be' stationary.
  - The white noise. (It is stationary.)
  - The SOI and # fish series.
- What about?
  - Daily sensex returns.
  - Random walk.
  
## Weakly stationary time series
A time series $\{x_t\}$ is said to be weakly stationary if

- If mean value $\mu_t$ is independent of $t$.
- If $\rho(s, t) = \rho(|s - t|)$ for all $s, t$.

A widely accepted convention is to call a weakly stationary time series as
stationary. 

Note that $\rho(s, t) = \rho(|s - t|) \Rightarrow \rho(s, s+h) = \rho(h)$.

## Descriptive statistics of time series - 1
```{r}
par(mfrow = c(3, 1))
acf(soi, 48, main = "SOI")
acf(rec, 48, main = "# fish")
ccf(soi, rec, 48, main = "SOI v. # fish", ylab = "CCF")
```

## Descriptive statistics of time series - 2
Key observations

- The '# fish' series appears to be lagging behind the SOI series. When SOI peaks
we can expect the # fish to peak 'after some time'.
- The CCF plot corroborates this belief. There is a strong CCF signal at $h = -6$.
It seems like the # fish respond to SOI with a lag of $6$ months.
  - Alternatively, # fish at time $t$ are strongly correlated with SOI level at
  time $t - 6$.
  - Further a negative value of CCF indicates that a rise in SOI level leads to
  a drop in the # fish.
- Such effects are fairly common in economic analysis. A sagging industrial
production might foretell that the economy is going to slow down a few months
down the line.

## Spectral analysis - 1
- A reasonably smooth (I choose to be non-technical) function can be expressed
as a sum of sines and cosines. 
- If $x(t)$ is the time series then
$$
x(t) = \sum_{\nu=-\infty}^\infty c_\nu \exp\left(i\nu t\right)
$$
- The 'amplitude' of a sinusoidal wave is $c_n$ given by
$$
c_\nu = \frac{1}{2\pi}\int_{-\pi}^{\pi}x(t)\exp\left(-i\nu t\right)dt.
$$
- The magnitude of $c_\nu$ indicates the dominance of the wave of frequency $\nu$.
- A plot of $|c_n|^2$ versus $n$ is called the spectrum or the periodogram.

## Spectral analysis - 2
Sunspots are known to peak every $11$ years. A time series of the number of
sunspots seen from $1700$ to $1988$ is
```{r}
plot(sunspot.year, ylab = "# sunspots", main = "Yearly sunspots")
```

## Spectral analysis - 3
A periodogram of the yearly sunspots shows highest power at $n = 1/11.1 = 0.0909$.
```{r}
mvspec(sunspot.year, log = "no", main = "Periodogram of yearly sunspots")
abline(v = 1/11.1, lty = 2, col = "red")
```

## Spectral analysis - 4
The southern oscillations are known to be quasi-periodic with a frequency betwee
$4$ and $7$ years.
```{r}
mvspec(soi, log = "no", main = "Periodogram of SOI")
abline(v = 0.2, lty = 2, col = "red")
```

## Spectral analysis - 5
A periodogram of the # fish shows a dominant cycle of $1$ year and another one of
$4$ years.
```{r}
mvspec(rec, log = "no", main = "Periodogram of # fish")
abline(v = 0.25, lty = 2, col = "red")
```

## Spectral analysis - 6
This is the why white noise is called so.
```{r}
mvspec(rnorm(5000), log = "no", main = "Periodogram of white noise")
```

## What does spectral analysis give us?
- The dominant frequencies indicate seasonality.
- You can remove seasonal effects by transforming the time series through an
appropriate filter.
- If the time signal is very noisy, you can smoothen it by transforming the
time series through a low pass filter.
- Let's apply what we learnt to the sensex daily returns.

## Sensex daily returns - 1
The periodogram of daily returns is:
```{r}
sensex_master <- read.csv("sensex.csv")
delta_sensex <- diff(sensex_master$Close)
daily_returns <- delta_sensex / sensex_master$Close[seq_along(delta_sensex)]
mvspec(daily_returns, log = "no", main = "Periodogram of daily returns")
```

## Sensex daily returns - 2
- Looks similar to white noise yet it is clear that the higher frequencies
are present to a lesser extent.
- It is quite possible that the very high frequencies are the inherent fluctuations
in the returns. In order to check if there is anything of interest at a slightly
lower frequency, we must filter out the noise.
- We therefore pass the daily returns time series through an averaging filter.
```{r echo = FALSE}
smoothed_returns <- filter(daily_returns, rep(1/3,3))
```

## Sensex daily returns - 3
The periodogram of the smoothed daily returns time series is
```{r}
mvspec(smoothed_returns[!is.na(smoothed_returns)], main = "Smoothed daily returns")
```
